Classless Java: Tuning Java Interfaces

### 2015-12-06 paper/sections/Formalization.tex
revised S3 and S4

\section{Interaction of Interface Methods with Interface Composition}
%Before formalizing Classless Java, it is helpful to informally discuss the behaviour
%of methods in Java 8 interfaces.

>>>>

\section{Interaction of Interface Methods with Interface Composition}
Before formalizing Classless Java and object interfaces, it is helpful
to informally discuss the behaviour of methods in Java 8
interfaces. In particular it is useful to understand how Java 8
interfaces differ from conventional trait models.

### 2015-12-06 paper/sections/Overview.tex
Overview/Intro
revised S3 and S4

\begin{figure}
\begin{tabular}{c|c}
\includegraphics[height=3.5cm]{PegasusDetail.pdf}\hspace{20pt} &
\begin{minipage}{7cm}
\vspace{-90pt}
\lstinputlisting[linerange=6-15]{../UseMixinLombok/src/pegasusJava/Main.java}% APPLY:linerange=PEGASUS_JAVA
%basicstyle=\ttfamily\scriptsize
\end{minipage}
\end{tabular}
\caption{The animal system (left: complete structure, right: code for simplified animal system).}\label{fig:pegasus}
\end{figure}

This section illustrates how the \mixinAnn{} annotation enables expressive
idioms using multiple inheritance. 
To propose a standard example for multiple inheritance, we show \Q@Animal@s with a two dimensional \Q@Point2D@ representing their \Q@location@.
Some kinds of animals are \Q@Horse@s and \Q@Bird@s.
Birds can \Q@fly@, thus their location need to be a three dimensional \Q@Point3D@.
Finally, we model \Q@Pegasus@ (one of the best known creatures in Greek
mythology) as a kind of \Q@Animal@ with the skills
of both \Q@Horse@s and \Q@Bird@s. A simple class diagram illustrating
the basic system in given in the left side of Figure~\ref{fig:pegasus}.\footnote{
Some research argues in favour of using subtyping for modelling taxonomies, other research argue against this practice, we do not wish to take sides in this argument, but to provide an engaging example.}

@@

\noindent There is a some annoyances here. Firstly, the sole
purpose of classes is to provide a way to instantiate
objects. Although (in this case) it takes only one line code to
provide each of those classes, this code is essentially boilerplate
code, which does not add behaviour to the system. Secondly, 
the namespace gets filled with three additional types. For example, 
both \texttt{Horse} and \texttt{HorseImpl} are needed: \texttt{Horse}
is needed because it needs to be an inferface so that \texttt{Pegasus}
can use multiple inheritance; and \texttt{HorseImpl} is needed to
provide object instantiation. 
Note that, for this very simple animal system, plain Java 8 anonymous
classes can be used to avoid these problems.  We could have simply
instantiated Pegasus using:

@@

\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int _x, int _y){return new Point2D(){
    int x=_x; int y=_y;
    public int x(){return x;}   public int y(){return y;}
    Point2D withX(int val){ return Point2D.of(val,this.y());}    
    Point2D withY(int val){ return Point2D.of(this.x(),val);}  
    };}
  Point2D withX(int val);    Point2D withY(int val);}
\end{lstlisting}
\end{comment}
Moreover, if the programmer wants a different implementation, he may 
provide an alternative implementation using \Q@default@ methods:
\begin{lstlisting}
@Obj interface Point2D{ 
  int x(); int y();
  default Point2D withX(int val){ ... };    default Point2D withY(int val){ ... }}
\end{lstlisting}

%\begin{comment}
\marco{I re-enabled this code, I think is needed for understandability}
is expanded into
\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int _x, int _y){return new Point2D(){
    int x=_x;    int y=_y;
    public int x(){return x;}    public int y(){return y;}
    };}
  default Point2D withX(int val){ ... }    default Point2D withY(int val){ ... }}
\end{lstlisting}

\noindent We can provide implementation for methods; 
we generate code only for methods needing implementation. Thus, programmers can easily tailor the behaviour on their special needs.
%\end{comment}

@@

\paragraph{Summary} Dealing with state (mutable or not) in object interfaces
relies on a notion of abstract state, where only methods that interact with
state are available to users. Object interfaces provide support for four
different types of abstract state operations, which are summarized in
Figure~\ref{fig:abstractstate}.  The abstract state operations are determined by
naming conventions and the types of the methods. Fluent setters are a variant of
(void) setters, and are discussed in detail in Section~\ref{sec:dsls}.

@@

We are tempted to assume that a language extension is needed. 
%Instead, with object interfaces, another approach is possible
Instead, with our language \emph{restriction}, another approach is possible;often in programming languages ``freedom is slavery''.

@@

\noindent \Q@Point3D@ includes a
\Q@with@ method, taking a \Q@Point2D@ as an argument.
Other wither methods (such as \Q@withX@) functionally update a field at a time.  This can be
inefficient, and sometimes hard to maintain.  Often we want to update multiple
fields simultaneously, for example using another object as source.  Following
this idea, the method \Q@with(Point2D)@ is an example of a (functional)
properties updater: it takes a certain type of object and returns a copy of the
current object with all fields that match the fields of parameter type
updated. The idea is that the result should be like \Q@this@, but
modified to be as similar as possible to the parameter.

@@

\paragraph{Generated boilerplate}
Just to give a feeling on how much boring code \mixin is generating, we show the
generated code for the \texttt{Point3D} in Figure~\ref{fig:boilerplate}. The
generated code is very mechanical, writing such code by hand is error-prone. For
example a distracted programmer may swap the arguments of calls to
\Q@Point3D.of@.  Note how \Q@with-@ methods are automatically refined in their
return type, so that code like:

>>>>

\begin{figure}
\begin{tabular}{c|c}
\includegraphics[height=3.5cm]{PegasusDetail.pdf}\hspace{20pt} &
\begin{minipage}{7cm}
\vspace{-90pt}
\lstinputlisting[linerange=6-15]{../UseMixinLombok/src/pegasusJava/Main.java}% APPLY:linerange=PEGASUS_JAVA
%basicstyle=\ttfamily\scriptsize
\end{minipage}
\end{tabular}
\caption{The animal system (left: complete structure, right: code for simplified animal system).}\label{fig:pegasus}
\end{figure}
\bruno{Yanlin: We need space, but we cannot make the figure {\bf unreadable}; the
  diagram fonts are too small now. Please increase the size of the
  picture a bit. You can also use more space in the Java code. Since
  you are on it please fix Point2D: there's no reason for the space
  between x and y.}

This section illustrates how object interfaces, expressed using the \mixinAnn{} annotation, enables powerful
idioms using multiple inheritance in Java. 
To propose a standard example for multiple inheritance, we show \Q@Animal@s with a two dimensional \Q@Point2D@ representing their \Q@location@.
Some kinds of animals are \Q@Horse@s and \Q@Bird@s.
Birds can \Q@fly@, thus their location need to be a three dimensional \Q@Point3D@.
Finally, we model \Q@Pegasus@ (one of the best known creatures in Greek
mythology) as a kind of \Q@Animal@ with the skills
of both \Q@Horse@s and \Q@Bird@s. A simple class diagram illustrating
the basic system in given in the left side of Figure~\ref{fig:pegasus}.\footnote{
Some research argues in favour of using subtyping for modelling taxonomies, other research argues against this practice, we do not wish to take sides in this argument, but to provide an engaging example.}

@@

\noindent There are some annoyances here. Firstly, the sole
purpose of classes is to provide a way to instantiate
objects. Although (in this case) it takes only one line code to
provide each of those classes, this code is essentially boilerplate
code, which does not add behaviour to the system. Secondly, 
the namespace gets filled with three additional types. For example, 
both \texttt{Horse} and \texttt{HorseImpl} are needed: \texttt{Horse}
is needed because it needs to be an inferface so that \texttt{Pegasus}
can use multiple inheritance; and \texttt{HorseImpl} is needed to
provide object instantiation. 
Note that, for this very simple animal system, plain Java 8 anonymous
classes can be used to avoid these problems.  We could have simply
instantiated Pegasus using:

@@

\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int _x, int _y){return new Point2D(){
    int x=_x; int y=_y;
    public int x(){return x;}   public int y(){return y;}
    Point2D withX(int val){ return Point2D.of(val,this.y());}    
    Point2D withY(int val){ return Point2D.of(this.x(),val);}  
    };}
  Point2D withX(int val);    Point2D withY(int val);}
\end{lstlisting}
\end{comment}
Moreover, if the programmer wants a different implementation, he may 
provide an alternative implementation using \Q@default@ methods. For example:
\begin{lstlisting}
@Obj interface Point2D{ 
  int x(); int y();
  default Point2D withX(int val){ ... };    default Point2D withY(int val){ ... }}
\end{lstlisting}

%\begin{comment}
%\marco{I re-enabled this code, I think is needed for understandability}
\noindent is expanded into
\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int _x, int _y){return new Point2D(){
    int x=_x;    int y=_y;
    public int x(){return x;}    public int y(){return y;}
    };}
  default Point2D withX(int val){ ... }    default Point2D withY(int val){ ... }}
\end{lstlisting}

\noindent If an implementation is provided for methods,
only code for methods needing implementation is generated. Thus,
programmers can easily tailor the behaviour for their special needs.
%\end{comment}

@@

\paragraph{Summary} Dealing with state (mutable or not) in object interfaces
relies on a notion of abstract state, where only methods that interact with
state are available to users. Object interfaces provide support for four
different types of abstract state operations, which are summarized in
Figure~\ref{fig:abstractstate}.  The abstract state operations are determined by
naming conventions and the types of the methods. Fluent setters are a variant of
conventional setters, and are discussed in detail in Section~\ref{sec:dsls}.

@@

We may tempted to assume that a language extension is needed. 
%Instead, with object interfaces, another approach is possible
Instead, the \emph{restriction} of (object) interfaces to have no
fields, enlighten us that another approach is possible; often in programming languages ``freedom is slavery''.

@@

\noindent \Q@Point3D@ includes a
\Q@with@ method, taking a \Q@Point2D@ as an argument.
Other wither methods (such as \Q@withX@) functionally update a field at a time.  This can be
inefficient, and sometimes hard to maintain.  Often we want to update multiple
fields simultaneously, for example using another object as source.  Following
this idea, the method \Q@with(Point2D)@ is an example of a (functional)
properties updater: it takes a certain type of object and returns a copy of the
current object with all fields that match the fields of updated
parameter type. The idea is that the result should be like \Q@this@, but
modified to be as similar as possible to the parameter.

@@

\paragraph{Generated boilerplate}
Just to give a feeling on how much boring code \mixin is generating, we show the
generated code for the \texttt{Point3D} in Figure~\ref{fig:boilerplate}. The
generated code is mechanical. Writing such code by hand is error-prone. For
example a distracted programmer may swap the arguments of calls to
\Q@Point3D.of@.  Note how \Q@with-@ methods are automatically refined in their
return type, so that code like:
